The Space of Feasible Execution Times for Asynchronous Periodic Task Systems using Definitive Idle Times

### 2013-07-30 dit-paper_ieee.tex
changes by LG

\algsetup{
  indent=1em,
  linenosize=\small,
  linenodelimiter=\
}

@@

\begin{document}
\title{Describing the C-Space of Asynchronous Periodic Task Systems using
Definitive Idle Times}

@@

\begin{abstract}
Sensitivity analysis for real-time systems asks for efficient methods to determine feasibility
after small changes of execution times, such as when porting a system to a different platform.
The C-space, presented in \cite{bini2004schedulability,george2009characterization}, describes the
set of execution times for which a system is feasible. This paper extends previous works to
generalize the description of the C-space for Earliest Deadline First scheduling to asynchronous
systems and non-pairwise coprime periods. The feasibility gain offered by offsets is then
expressed as the volume ratio between the C-space of an asynchronous system and its corresponding
synchronous system.
\end{abstract}

@@

	Abstract models for real-time systems often require the determination of the
	(worst-case) computation time of a job (WCET), a value highly dependent on the
	platform on which the application is deployed.

@@

	In this section the current description of the C-space of synchronous
	constrained systems, which has been mostly covered by \cite{george2009characterization}, is reviewed.

	\subsection{Model}

		We consider a discrete timeline and tasks systems $\tau$ comprised of periodic concrete
		tasks with constrained deadlines on uniprocessor platforms. Each task
		$\tau_i$ is represented by a tuple $(O_i, C_i, D_i, T_i)$ where $O_i$ is the offset value,
		$C_i$ the WCET, $D_i$ the relative deadline and $T_i$ the period.
		The constrained deadline property implies that $D_i \leq T_i \; \forall i$. In
		the following, we differentiate between \emph{synchronous systems} (where $O_i
		= 0 \; \forall i$) and \emph{asynchronous systems} (which are not synchronous).

		For those systems, the \emph{Earliest Deadline First} (EDF) algorithm, in
		which jobs are prioritized according to their absolute deadline, has been
		shown to be optimal \cite{liu1973scheduling}, which means it
		correctly schedules any feasible system within that group.

		Furthermore, the following notations are used:
		\begin{itemize}
			\item $\tau = \{\tau_0, \cdots, \tau_{n-1}\}$
			\item $H = \operatorname{lcm}(T_0, \cdots, T_{n-1})$
			\item $O_{max} = \max (O_0, \cdots, O_{n-1})$
			\item An \emph{idle time} is an instant $t$ at which each job that arrived
			\emph{strictly} before $t$ has finished its execution.
			\item A \emph{busy period} is a time interval during which there is always
			at least one unfinished job available in the system and maximal
			with this property. Any busy period starts and ends
			with an idle time.
			\item The interval between the initial time and the first
			idle time is called the \emph{first busy period}.
		\end{itemize}

@@

	We recall the necessary and sufficient condition explained in \cite{baruah1999generalized, baruah1990algorithms,pellizzoni2005feasibility}, based on the demand-bound function.

	\begin{definition}
		The \textbf{demand-bound function (DBF)}
		defined for a task set $\tau$ and noted $\dbf{t_1, t_2}$, is equal to
		the maximal cumulated execution time of jobs of $\tau$ contained in the
		closed interval $[t_1, t_2]$.
	\end{definition}

@@

	The values of the $n_i$ are given by
	\[
		n_i(t_1, t_2) =
		\left\llbracket
			\left\lfloor
				\frac{t_2 - O_i - D_i}{T_i}
			\right\rfloor -
			\left\lceil
				\frac{t_1 - O_i}{T_i}
			\right\rceil + 1
		\right\rrbracket_0
	\]
	where $\llbracket x \rrbracket_0$ stands for $\max \{ 0, x \}$.

	A necessary and sufficient condition of feasibility can easily be written
	based on the DBF:

@@

	\subsection{C-space of synchronous systems}

		\subsubsection{Definition}

			The authors of \cite{george2009characterization} give the following definition of the C-space (based on \cite{bini2004schedulability}):
			\begin{definition}
				The \textbf{C-space} of a task system $\tau$ is a region of $n$ dimensions (where each dimension denotes the possible $C_i$ of a task of $\tau$) such that for any vector $C = \{ C_0, \cdots, C_{n-1}\}$ in it, $\tau$ is feasible.
			\end{definition}

    \subsubsection{Description using DBF}
      \label{sct:cspaceDescr}

@@

  \subsection{Our work}
    In this paper, we extend the work of \cite{george2009characterization} to the
    asynchronous case. In section \ref{sct:asyncDIT}, we present an extension of
    the concept of DIT adapted to the asynchronous case called first periodic DIT
    ($t_d$) along with an algorithm to determine its existence and value for any system,
    then we show the influence of some factors on its existence. In Section
    \ref{sct:asyncCspace}, we show how to describe the C-space of an asynchronous
    system based on the DBF test, we show why the interval $[t_d, t_d + H]$ is
    sometimes sufficient to consider and we give a numerical example.
    Finally, in section \ref{sct:expCspaceGain}, we use a simulation to quantify the feasibility
    gain of asynchronous system in terms of volume of the C-space.

@@

	\subsection{Solving the modular system}
		The existence conditions of	Section~\ref{sct:FPDITexist} can directly be used to calculate the value of
		the FPDIT (if there is one), because for a given set of $a_i$ values they are
		very similar to an instance of the Chinese Remainder Problem (CRP), which is
		easily solved using Gauss's CRP solving method for non-coprime moduli
		\cite{gauss1965disquisitiones}.

		However this is very inefficient as there are exponentially many different
		combinations of $a_i$ values in the number of tasks, and resolving
		exponentially many CRP is prohibitively slow. A better method for finding the
		FPDIT consists in splitting Gauss's algorithm into two separate parts, as
		shown in Algorithm~\ref{alg:multiACRP}.

		\begin{algorithm}
			\caption{Modified Gauss Algorithm}
			\label{alg:multiACRP}
			\begin{algorithmic}[1]
				\REQUIRE $A$ is a list of the sets of possible $a_i$ per task
				\REQUIRE $H$ is the hyperperiod of the system

				\STATE $sumchunks \leftarrow \text{empty list}$
				\FORALL{$\tau_j \in \tau$}
					\STATE $taskTerms \leftarrow \text{empty list}$ 
					\STATE $Mj \leftarrow \frac{H}{T_j}$
					\STATE $invMj \leftarrow Mj^{-1} (mod \; T_j)$
					\FORALL{$a_i \in A(\tau_j)$}
						\STATE $\text{append} \; (a_i \cdot Mj \cdot invMj) \; \text{to} \; taskTerms$
					\ENDFOR
					\STATE $\text{append} \; taskTerms \; \text{to} \; sumchunks$
				\ENDFOR

				\STATE $results \leftarrow \text{empty list}$
				\STATE append $0$ to $results$
				\FORALL{$\tau_j \in \tau$}
					\STATE $newresults \leftarrow \text{empty list}$
					\FORALL{$a_i \in A(\tau_j)$}
						\FORALL{$r \in results$}
							\STATE $\text{append} \; (r + sumchunks(\tau_j,i)) \; \text{to} \; newresults$
						\ENDFOR
					\ENDFOR
					\STATE $results \leftarrow newresults$
				\ENDFOR
				\RETURN $r \; (mod \; H)$ for all $r \in results$
			\end{algorithmic}
		\end{algorithm}

		The global structure of the original algorithm is simply a sum of factors, each term being dependent on which $a_i$ value is chosen. A major
		speedup is achieved by first computing all possible terms for each task
		then creating all possible DIT values by summing the terms in all possible
		ways. The number of results is still exponential in the number of tasks, but
		the only operation that is repeated exponentially many times is now a sum.
		This allows us to profit from the high chance of having DITs when the CDF
		factor is close to one, but with only a slight increase in computation time as the sizes of the $[D_i, T_i]$ intervals grow.

@@

The first DIT happens at time $t_d = 7$. The intervals to consider are thus of the form $[0, t]$ where $t$ is a deadline happening before or at time $t_d$. This yields 2 intervals ($[0, 2]$ and $[0, 7]$) and the following constraints describing the C-space, none of them being redundant:
$$
\left\{
  \begin{array}{ccccc}
    & & C_2 & \leqslant & 2 \\
    C_1 & + & 2 C_2 & \leqslant & 7
  \end{array}
\right.
$$
This time, the C-space only contains 8 points: $(1, 1)$, $(2, 1)$, $(3, 1)$, $(4, 1)$, $(5, 1)$, $(1, 2)$, $(2, 2)$, $(3, 2)$.

Fig. \ref{fig:cspaceComp} shows the two C-spaces in the plane. The dark grey region is the C-space of the synchronous system and the light grey region is the surface gain given by the asynchronous system.

@@

Future works could also study the shape of the C-space and how it is affected by some factors. An interesting path would be the effect of the CDF on the number of constraints describing the C-space, as we know that in the implicit case this description is simple ($U_{tot} \leqslant 1$). While we quantified the feasibility gain by looking at the ratio between sizes of C-space, another approach would be to look at the ratio between the minimal distance of the pareto front of the C-space, which would give a lower bound of the gain given by the offsets.

>>>>

\algsetup{
  indent=2em,
  linenosize=\small,
  linenodelimiter=\
}

@@

\begin{document}
\title{The Space of Feasible Execution Times for Asynchronous Periodic Task Systems using
Definitive Idle Times}

@@

\begin{abstract}
Sensitivity analysis for real-time systems provides for efficient methods to determine feasibility
conditions  in the case of small changes in task parameters, such as when porting a system to a
different platform. The C-space, presented in \cite{bini2004schedulability,george2009characterization}, describes the
set of execution times for which a system is feasible. This paper extends previous works to
generalize the description of the C-space for Earliest Deadline First scheduling to
systems with initial offset. The feasibility gain offered by offsets is then
expressed as the volume ratio between the C-space of an asynchronous system and its corresponding
synchronous system.
\end{abstract}

@@

	Abstract models for real-time systems often require the determination of the
	(worst-case) execution time of a job (WCET), a value highly dependent on the
	platform on which the application is deployed.

	Feasibility conditions...

@@

    In section~\ref{sct:relatedWorks}, the description of the C-space of synchronous
	constrained systems, which has been mostly covered by \cite{george2009characterization}, is revisited.
	This description is based on the DBF feasibility condition described in
	\cite{baruah1999generalized,baruah1990algorithms} and is found with exponential complexity using
	the notion of definitive idle times (DIT).

	In section~\ref{sct:asyncDIT}, we present an extension of the concept of DIT adapted to the
	asynchronous case called first periodic DIT
    ($t_d$) along with an algorithm to determine its existence and value for any system,
    then we show the influence of some factors on its existence.

    In Section~\ref{sct:asyncCspace}, we show how to describe the C-space of an asynchronous
    system based on the DBF test, we show why the interval $[t_d, t_d + H]$ is
    sometimes sufficient to consider and we give a numerical example.

    Finally, in section~\ref{sct:expCspaceGain}, we use a simulation to quantify the feasibility
    gain of asynchronous system in terms of volume of the C-space.

\section{Related works}
\label{sct:relatedWorks}
	\subsection{Model}

		We consider a discrete timeline and tasks systems $\tau$ comprised of periodic concrete
		tasks with constrained deadlines on uniprocessor platforms. Each task
		$\tau_i$ is represented by a tuple $(O_i, C_i, D_i, T_i)$ where $O_i$ is the offset value,
		$C_i$ the WCET, $D_i$ the relative deadline and $T_i$ the period.
		The constrained deadline property implies that $\forall i,\; D_i \leq T_i$. In
		the following, we differentiate between \emph{synchronous systems} (where $O_i
		= 0 \; \forall i$) and \emph{asynchronous systems} (which are not synchronous).

		For those systems, the preemptive \emph{Earliest Deadline First} (EDF) algorithm, in
		which jobs are prioritized according to their absolute deadline, has been
		shown to be optimal \cite{liu1973scheduling}, which means it
		correctly schedules any feasible system within that group.

		Furthermore, the following notations are used:
		\begin{itemize}
			\item $\tau = \{\tau_1, \cdots, \tau_{n}\}$
			\item $H = \operatorname{lcm}(T_1, \cdots, T_{n})$
			\item $O_{max} = \max (O_1, \cdots, O_{n})$
			\item An \emph{idle time} is an instant $t$ at which each job that arrived
			\emph{strictly} before $t$ has finished its execution.
			\item A \emph{busy period} is a time interval between two consecutive idle times in
			which at least one job is executing.
			\item The interval between the initial time ($t=0$) and the first
			idle time is called the \emph{first busy period}.
		\end{itemize}

@@

	We recall the necessary and sufficient condition explained in \cite{baruah1999generalized, baruah1990algorithms}, based on the demand-bound function.

	\begin{definition}
		The \textbf{demand-bound function (DBF)}
		defined for a task set $\tau$ in a time interval $[t_1, t_2]$ and denoted $\dbf{t_1, t_2}$, is
		equal to the maximal cumulated execution time of jobs of $\tau$ contained in the
		closed interval $[t_1, t_2]$.
	\end{definition}

@@

	The values of the $n_i$ are given by \cite{baruah1999generalized}
	\[
		n_i(t_1, t_2) =
		\left\llbracket
			\left\lfloor
				\frac{t_2 - O_i - D_i}{T_i}
			\right\rfloor -
			\left\lceil
				\frac{t_1 - O_i}{T_i}
			\right\rceil + 1
		\right\rrbracket_0
	\]
	where $\llbracket x \rrbracket_0$ stands for $\max \{ 0, x \}$.

	A necessary and sufficient condition of feasibility can easily be written
	based on the DBF \cite{baruah1999generalized}

@@

	\subsection{C-space of synchronous systems}
	\label{sct:cspaceDescr}

			The authors of \cite{george2009characterization} give the following definition of the C-space (based on \cite{bini2004schedulability}):
			\begin{definition}
				The \textbf{C-space} of a task system $\tau$ is a region of $n$ dimensions (where each dimension denotes the possible $C_i$ of a task of $\tau$) such that for any vector $C = \{ C_1, \cdots, C_{n}\}$ in it, $\tau$ is feasible.
			\end{definition}

@@

@@

	\subsection{Solving the modular system}
		Finding the FPDIT of a system is similar to finding at least one synchronous
		instant (an instant at which all tasks release a job simultaneously) out of a
		collection of systems with different offsets. The existence conditions can
		directly be used to calculate the value of the FPDIT (if there is one),
		because for a given set of $a_i$ values they are very similar to an instance
		of the Chinese Remainder Problem (CRP), which is easily solved using Gauss's
		CRP solving method for non-coprime $T_i$ \cite{gauss1965disquisitiones}.

		However this is very inefficient as there are exponentially many different
		combinations of $a_i$ values in the number of tasks. A better method for finding the FPDIT consists in splitting Gauss's
		algorithm into two separate parts. Its global structure is simply a sum of
		factors, each term being dependent on which $a_i$ value is chosen. A major speedup can
		be achieved by first computing all possible terms for each task then creating all
		possible DIT values by summing the terms in all possible ways. The number of results is still
		exponential in the number of tasks, but the only operation that is repeated
		exponentially many times is now a sum.

@@

The first DIT happens at time $t_d = 7$. The intervals to consider are thus of the form $[0, t]$ where $t$ is a deadline happening before or at time $t_d$. This yields two intervals ($[0, 2]$ and $[0, 7]$) and the following constraints describing the C-space, none of them being redundant:
$$
\left\{
  \begin{array}{ccccc}
    & & C_2 & \leqslant & 2 \\
    C_1 & + & 2 C_2 & \leqslant & 7
  \end{array}
\right.
$$
This time, the C-space only contains 8 points: $(1, 1)$, $(2, 1)$, $(3, 1)$, $(4, 1)$, $(5, 1)$, $(1, 2)$, $(2, 2)$, $(3, 2)$.

Fig. \ref{fig:cspaceComp} shows the two C-spaces in the plane. The dark gray region is the C-space of the synchronous system and the light gray region is the surface gain given by the asynchronous system.

@@

Future works could also study the shape of the C-space and how it is affected by some factors. An
interesting path would be the effect of the CDF on the number of constraints describing the C-space,
as we know that in the implicit case this description is simple ($U_{tot} \leqslant 1$). While we
quantified the feasibility gain by looking at the ratio between sizes of C-space, another approach
would be to look at the ratio between the minimal distance of the border of the C-space, which would
give a lower bound of the gain given by the offsets.
